How to Manage My Data? With Machine-Interpretable GDPR Rights!

### 2024-09-15 paper.tex
finalise conclusions + add abstract

% PLEASE USE THIS FILE AS A TEMPLATE FOR THE PUBLICATION 
% Check file IOS-Book-Article.tex
%

\documentclass{IOS-Book-Article}     %[seceqn,secfloat,secthm]
%\usepackage{mathptmx}
%\usepackage[T1]{fontenc}
%\usepackage{times}%
\usepackage{graphicx}
\usepackage{listings}
\usepackage{minted}
\setminted{frame=lines,
    framesep=1mm,
    fontsize=\fontsize{7}{8},
    breaklines=true,
    linenos
    }
 %% auto break lines   
\lstset{breaklines=true}

@@

\begin{document}
\begin{frontmatter}          % The preamble begins here.
%
%\pretitle{}
\title{Delete My Data! Making GDPR Rights Machine-Interpretable and Exercisable}
\runningtitle{I Know My Rights! Making GDPR Rights Machine-Interpretable and Exercisable}
%\subtitle{Making GDPR Rights Machine-Interpretable and Exercisable}

\author[A]{\fnms{Beatriz} \snm{Esteves}
\thanks{Corresponding Author: Beatriz Esteves, \email{beatriz.esteves@ugent.be}.}},
\author[B]{\fnms{Harshvardhan J.} \snm{Pandit}},
\author[C]{\fnms{Georg P.} \snm{Krog}},
\author[D]{\fnms{Paul} \snm{Ryan}}
\runningauthor{B. Esteves et al.}
\address[A]{IDLab, Ghent University -- imec, Ghent, Belgium}
\address[B]{ADAPT Centre, Dublin City University, Dublin, Ireland}
\address[C]{Signatu AS, Oslo, Norway}
\address[D]{ADAPT Centre, Dublin City University, and Uniphar PLC, Dublin, Ireland}

\begin{abstract}
abstract
\end{abstract}

@@

The General Data Protection Regulation (GDPR)~\cite{gdpr} grants data subjects a set of rights designed to protect their personal data and to ensure that they have greater control over how their data is collected, processed, and used by organizations.
These rights include the right to access personal data, the right to rectification of inaccurate data, the right to erasure (commonly known as the ``right to be forgotten''), and the right to data portability, which allows individuals to ask for their data to be transferred between data controllers.
Additionally, the GDPR provides rights to restrict processing, object to data use, and avoid automated decision-making.
These rights, combined with the transparency and accountability measures imposed on organizations, aim to strike a balance between the interests of data subjects and the legitimate needs of businesses and institutions in the digital age.
However, despite the comprehensive rights granted to individuals under regulations such as the GDPR, a `technological gap', i.e., a significant lack of efficient tools, for exercising these rights remains~\cite{bernes_enhancing_2022}.
% some citations for the sentences below would be nice
Many organizations struggle to provide accessible, user-friendly mechanisms for individuals to manage their personal data, often relying on cumbersome manual processes, which can lead to delays, or even non-compliance.
For data subjects, the process of requesting access, rectification, or deletion of their data can be time-consuming, unclear, and inconsistent across different platforms.
Additionally, there is a lack of standardisation in how organizations handle these requests, further complicating the process.
Without more automated and streamlined tools, individuals face unnecessary barriers to effectively controlling their data, undermining the intent of regulations designed to empower them.

In this context, we present a specification to express information regarding rights exercise and management in a machine-interpretable format, using semantic Web standards such as RDF.
In doing so, we begin to address the automation, interoperability and standardisation challenges identified above, towards enabling the development of tools for assisting individuals and organisations in right exercising activities. 
Towards achieving this goal, we reuse and extend the Data Privacy Vocabulary~(DPV)~\cite{pandit2024dpv}, which allows the expression of information related to legislative requirements such as the GDPR.
The developed extension provides a taxonomy to enable representing justifications associated with the non-fulfilment, non-requirement, delays, and exercising of processes, including rights exercising.
Beyond DPV, the contributions in this article also promote the usage of the Open Digital Rights Language (ODRL)~\cite{iannella_odrl_2018}, Data Catalog (DCAT)~\cite{albertoni_dcat_2024}, DCMI Metadata~\cite{dcmi_2020}, and provenance (PROV)~\cite{lebo_prov_2013} ontologies to express information about 
(i) linking personal data processing activities to applicable rights,
(ii) providing notices related to said rights,
(iii) documenting the exercise of rights, and
(iv) GDPR rights requests.
The resulting specification is being developed in the context of the W3C Data Privacy Vocabularies and Controls Community Group (DPVCG).

@@

Even though the existence of these rights has been brought forward has a huge step for individuals empowerment over their personal data, no standards have been defined for the management of information related to rights exercising.
An exception to this statement might be the `Data Rights Protocol'\footnote{\url{https://datarightsprotocol.org}}, developed by Consumer Reports Innovation Lab\footnote{\url{https://innovation.consumerreports.org}}, which ``defines a web protocol encoding a set of standardized request/response data flows such that Users can exercise Personal Data Rights provided under regulations like the California Consumer Privacy Act, General Data Protection Regulation, and other regulatory or voluntary bases, and receive affirmative responses in standardized formats''\footnote{From \url{https://github.com/consumer-reports-innovation-lab/data-rights-protocol}}.
It does not however cover all GDPR rights and does not promote the usage of standard vocabularies to record information about rights being exercised.
Since semantic technologies promote interoperability, allowing information and its interpretability to be consistently defined across different platforms, improving the discoverability, management, and preservation of resources, we will promote the usage of semantic Web standards for the definition of a specification to express rights exercisement and management.
As such, DPV\footnote{Preferred namespace and prefix are \url{https://w3id.org/dpv\#} and \texttt{dpv}.}\cite{pandit2024dpv} is the main driver of the specification presented in this work, as it is the most comprehensive data protection-related vocabulary, providing a set of taxonomies for the expression of metadata related to the usage of personal data aligned with legal requirements, e.g., the GDPR.
Additionally, the usage of 
ODRL\footnote{Preferred namespace and prefix are \url{http://www.w3.org/ns/odrl/2/} and \texttt{odrl}.}\cite{iannella_odrl_2018}, 
DCAT\footnote{Preferred namespace and prefix are \url{http://www.w3.org/ns/dcat\#} and \texttt{dcat}.}\cite{albertoni_dcat_2024}, and 
PROV-O\footnote{Preferred namespace and prefix are \url{http://www.w3.org/ns/prov\#} and \texttt{prov}.}\cite{lebo_prov_2013}, 
which are Web standards developed and maintained by the W3C, is also promoted to express information about policies, catalogs of resources and activity provenance information, respectively.
The DCMI Metadata Terms standard\footnote{Preferred namespace and prefix are \url{http://purl.org/dc/terms/} and \texttt{dcterms}.}\cite{dcmi_2020}, recommended to be used for metadata-keeping by other other Web standards, including ODRL and DCAT, is also reused.

@@

Based on the GDPR provisions for data subject rights identified in Section~\ref{sec:sota}, the following requirements are addressed in the proposed specification:
\begin{inlineabc}
    \item[(\textit{Section~\ref{sec:applicable-rights}})] information about the existence of rights related to personal data processing activities,
    \item[(\textit{Section~\ref{sec:justifications}})] justifications for the exercising of said rights,
    \item[(\textit{Section~\ref{sec:notices}})] notices related to the fulfilment or non-fulfilment of rights,
    \item[(\textit{Section~\ref{sec:records}})] records of rights-related activities, and
    \item[(\textit{Section~\ref{sec:policies}})] GDPR-related rights requests as machine-executable policies.
\end{inlineabc}
The work is publicly available\footnote{Accessible at  \url{https://besteves4.github.io/dpv-rights-exercising}.} and includes examples for each requirement identified above, which are omitted from this article due to size restrictions.

@@

Beyond jurisdiction-dependence, applicable data subject rights also depend on the legal ground used to process personal data.
In the case of the GDPR, DPV's GDPR extension\footnote{Preferred namespace and prefix are \url{https://w3id.org/dpv/legal/eu/gdpr\#} and \texttt{eu-gdpr}.} contains the terms to represent the rights available under GDPR, as well as a mapping of which rights are applicable based on the used legal basis.
As such, apart from detailing information about what personal data is being processed, how, where, by whom, and for what purpose, a \texttt{dpv:Process} can also be used to indicate applicable rights.
Data controllers can use this approach to express which rights apply, including those beyond GDPR, such as the EU’s fundamental rights and rights outlined in other EU regulations or jurisdictions, i.e., the \texttt{dpv:hasScope} property can be used to indicate applicable rights by jurisdiction.
Listing~\ref{lst:rights} highlights the modeling of a personal data process using DPV, where \texttt{dpv:hasScope} is used to express existing rights under EU's GDPR when data is processed based on consent and its similar UK-based counterpart.

\begin{listing}[ht]
\caption{Associate a process with applicable rights from different jurisdictions.}
\label{lst:rights}
\begin{minted}{turtle}
ex:ProcessEmailForServiceProvision a dpv:Process ;
    dpv:hasDataController ex:DataController ;
    dpv:hasPersonalData pd:EmailAddress ;
    dpv:hasProcessing dpv:Use ;
    dpv:hasPurpose dpv:ServiceProvision ;
    dpv:hasScope [
        dpv:hasLegalBasis eu-gdpr:A6-1-a ;
        dpv:hasJurisdiction loc:EU ;
        dpv:hasApplicableLaw legal-eu:law-GDPR ;
        dpv:hasRight eu-gdpr:A7-3, eu-gdpr:A13, eu-gdpr:A14, eu-gdpr:A15, eu-gdpr:A16, 
            eu-gdpr:A17, eu-gdpr:A18, eu-gdpr:A20, eu-gdpr:A22, eu-gdpr:A77 ] ;
    dpv:hasScope [
        dpv:hasLegalBasis ex:GB-GDPR-consent ;
        dpv:hasJurisdiction loc:GB ;
        dpv:hasApplicableLaw legal-gb:law-GDPR, legal-gb:law-DPA ;
        dpv:hasRight ex:RightOfAccess ] . # not complete, other rights exist

@@

As a result of the collection of requirements for GDPR rights exercising, a \texttt{Justification} taxonomy\footnote{Preferred namespace and prefix are \url{https://w3id.org/dpv/justifications\#} and \texttt{justifications}.}, to provide reasons or explanations related to the fulfilment, non-fulfilment, delay and exercise of rights, was developed.
Examples of justifications include individual's identity not being verifiable or a request being considered overly excessive or burdensome.
Justifications can also explain why certain processes are necessary or being delayed, such as the ground to exercise the right to object or requiring additional information to move forward with a process.
To facilitate the expression of these detailed justifications, this extension introduces concepts that extend DPV's \texttt{Justification} concept.
Moreover, since such concepts can be reused for other data protection-related activities, e.g., data breach reports, the concepts were defined in a generic way so that they can be reused in distinct contexts.

@@

Right exercise notices should inform data subjects about where and how to exercise an active right, what information is needed, or provide updates on a submitted right request.
DPV’s \texttt{isExercisedAt} property can be used to link rights to specific exercise points, such as through right exercise notices, and to connect the required information with a process using DPV’s \texttt{hasProcess} property. DPV's \texttt{hasRecipient}, \texttt{hasStatus}, and \texttt{hasJustification} properties can be employed to update the data subject on the status of their right exercising activity, including any justification for why the right is being exercised.
In addition to the properties mentioned earlier, DCMI's concepts can be reused to indicate temporal and other metadata information related to notices.
Table~\ref{tab:dcterms} summarises the terms reused from DCMI throughout this specification.
The notices should also provide information about the data controller, as well as the entity responsible for implementing the notice.
This could be the data controller or another entity acting on their behalf, represented using \texttt{dpv:hasDataController} and \texttt{dpv:isImplementedByEntity}, respectively. Additionally, \texttt{foaf:page}\footnote{FOAF's vocabulary namespace and prefix are \url{http://xmlns.com/foaf/0.1/} and \texttt{foaf}.} can be used to reference a document, such as a Webpage, that contains the full details of the notice and offers further functionalities.

\begin{table}[ht]
    \centering\fontsize{7}{8}
    \begin{tabular}{c|c}
        \textbf{\texttt{dcterms} Concepts} & \textbf{Usage} \\ \hline\hline
        \texttt{created},\texttt{issued},\texttt{modified} & Date of creation, issuance or modification of a notice \\ \hline
        \texttt{description} & Description of a notice \\ \hline
        \texttt{identifier} & Identifier of a notice \\ \hline
        \texttt{language} & Language in which the notice is provided \\ \hline
        \texttt{publisher} & Entity responsible for making the notice available \\ \hline
        \texttt{valid} & Date of validity of a notice \\
    \end{tabular}
    \caption{List of terms reused from the DCMI Metadata Terms vocabulary~\cite{dcmi_2020}.}
    \label{tab:dcterms}
\end{table}

@@

\begin{listing}[ht]
\caption{Notice of non-fulfillment related to an exercised right to erasure.}
\label{lst:notice}
\begin{minted}{turtle}
ex:RejectRightToErasure a dpv:RightNonFulfilmentNotice ;
    dcterms:issued "2024-09-06"^^xsd:date ;
    dcterms:description "Notice of non-fulfillment related to an exercised right to erasure" ;
    dcterms:identifier "x4ghyun-658393" ;
    dcterms:language "EN" ;
    dcterms:publisher ex:DataController ;
    dpv:hasRight eu-gdpr:A17 ;
    dpv:hasDataController ex:DataController ;
    dpv:isImplementedByEntity ex:DataController ;
    foaf:page <https://example.org/DataController/RejectRightToErasure> ;
    dpv:hasRecipient ex:DataSubject ;
    dpv:hasStatus dpv:RequestUnfulfilled ;
    dpv:hasJustification justifications:FreedomOfExpressionImpaired .

@@

\begin{listing}[ht]
\caption{Record maintained by the data controller for a data subject's rights-related requests.}
\label{lst:record}
\begin{minted}{turtle}
ex:catalog-001 a dpv:RightExerciseRecord, dcat:Catalog ;
    dcterms:description "Record maintained by the data controller for a data subject's rights-related requests" ;
    dcterms:publisher ex:DataController ;
    dcterms:created "2023-10-23T08:37:25"^^xsd:dateTime ;
    dcterms:modified "2023-11-03T08:37:25"^^xsd:dateTime ;
    dcat:record ex:DS-record ;
    dcat:catalog ex:request-001, ex:request-002 .

@@

\begin{figure}[ht]
\caption{Lifecycle of DPV's concepts to model the status of a request.}
\label{fig:status}
\centering
\includegraphics[width=0.7\textwidth]{img/request-status.png}
\end{figure}

@@

\begin{listing}[ht]
\caption{Right exercise activity where the data controller requires further action from the data subject to be able to proceed with the request.}
\label{lst:activity}
\begin{minted}{turtle}
ex:SARRequiresAction a dpv:RightExerciseActivity, prov:Activity ;
    dcterms:description "Data controller requires further action from data subject" ;
    dcterms:created "2023-11-02T16:09:21"^^xsd:dateTime ;
    prov:wasAssociatedWith ex:DataController ;
    dpv:hasDataSubject ex:DataSubject ;
    dpv:hasStatus dpv:RequestRequiresAction ;
    dpv:hasJustification justifications:IdentityVerificationFailure ;
    dpv:hasProcess [
        dpv:hasPersonalData pd:OfficialID ;
        dpv:hasProcessing dpv:MakeAvailable ;
        dpv:hasPurpose dpv:IdentityVerification ;
        dpv:hasRecipientDataController ex:DataController ;
        dpv:isImplementedByEntity ex:DataSubject ] .
\end{minted}
\end{listing}

@@

\begin{listing}[ht]
\caption{Request policy to exercise GDPR right to erasure.}
\label{lst:policy}
\begin{minted}{turtle}
ex:delete-request a odrl:Request ;
    odrl:uid "3456-7890-1234-5678-9012"^^xsd:string ;
    dcterms:description "Data subject requests data controller to delete their data." ;
    odrl:obligation ex:DS-delete-data ;
    odrl:obligation ex:DS-delete-notice .

@@

In this article, we proposed a semantic specification for representing information about rights exercisement and management based on the requirements of the GDPR.
The specification provides a machine-interpretable and executable model to represent information regarding data subject rights, concepts to justify their fulfilment or non-fulfilment, notices to communicate about them with other entities, records for auditing and policies for automated execution of responses to rights requests.
Through these, we hope to enable related tools and processes to handle obligations related to rights in an interoperable manner.

\bibliography{ref}
\bibliographystyle{vancouver}

>>>>

\documentclass{IOS-Book-Article}     %[seceqn,secfloat,secthm]
\usepackage{graphicx}
\usepackage{listings}
\usepackage{minted}
\setminted{frame=lines,
    framesep=1mm,
    fontsize=\fontsize{7}{8},
    breaklines=true,
    linenos
    }
 %% auto break lines   
\lstset{breaklines=true}

@@

\begin{document}
\pagestyle{plain}
\pagenumbering{arabic}
\begin{frontmatter}          % The preamble begins here.
% \title{Delete My Data! Making GDPR Rights Machine-Interpretable and Exercisable}
% \runningtitle{I Know My Rights! Making GDPR Rights Machine-Interpretable and Exercisable}
% \title{How to Control My Data? With Machine-Interpretable GDPR Rights!}
\title{How to Manage My Data? With Machine-Interpretable GDPR Rights!}

\author[A]{\fnms{Beatriz} \snm{Esteves}
\thanks{Corresponding Authors: Beatriz Esteves \email{beatriz.esteves@ugent.be} and Harshvardhan J. Pandit \email{me@harshp.com}}},
\author[B]{\fnms{Harshvardhan J.} \snm{Pandit}},
\author[C]{\fnms{Georg P.} \snm{Krog}},
\author[D]{\fnms{Paul} \snm{Ryan}}
\runningauthor{B. Esteves et al.}
\address[A]{IDLab, Ghent University -- imec, Ghent, Belgium}
\address[B]{ADAPT Centre, Dublin City University, Dublin, Ireland}
\address[C]{Signatu AS, Oslo, Norway}
\address[D]{ADAPT Centre, Dublin City University, and Uniphar PLC, Dublin, Ireland}

\begin{abstract}
The EU GDPR is a landmark regulation that introduced several rights for individual to obtain information and control how their personal data is being processed, as well as receive a copy of it. However, there are gaps in the effective use of rights due to each organisation developing custom methods for rights declaration and management, thereby burdening individuals. Simultaneously, there is a technological gap as there is no single consistent standards-based mechanism that can automate the handling of rights for both organisations and individuals.
In this article, we present a specification for exercising and managing rights in a machine-interpretable format based on semantic web standards.
Our approach uses the comprehensive Data Privacy Vocabulary (DPV) to create a streamlined workflow for individuals to understand what rights exist, how and where to exercise them, and for organisations to effectively manage them -- including in complex cases such as identity verification.
This work pushes the state of the art in GDPR rights management and is crucial for data reuse and rights management under technologically intensive developments, such as Data Spaces.
\end{abstract}

@@

The General Data Protection Regulation (GDPR)~\cite{gdpr} grants data subjects a set of rights designed to protect their personal data and to ensure that they have greater control over how their data is collected, processed, and used by organisations.
These rights include the right to access personal data, the right to rectification of inaccurate data, the right to erasure (commonly known as the ``right to be forgotten''), and the right to data portability, which allows individuals to ask for their data to be transferred between data controllers.
Additionally, the GDPR provides rights to restrict processing, object to data use, and avoid automated decision-making.
These rights, combined with the transparency and accountability measures imposed on organisations, aim to strike a balance between the interests of data subjects and the legitimate needs of businesses and institutions in the digital age.
However, despite the comprehensive rights granted to individuals under regulations such as the GDPR, a `technological gap', i.e., a significant lack of efficient tools, for exercising these rights remains~\cite{bernes_enhancing_2022}.
% some citations for the sentences below would be nice
Many organisations struggle to provide accessible, user-friendly mechanisms for individuals to manage their personal data, often relying on cumbersome manual processes, which can lead to delays, or even non-compliance.
For data subjects, the process of requesting access, rectification, or deletion of their data can be time-consuming, unclear, and inconsistent across different platforms.
Additionally, there is a lack of standardisation in how organisations handle these requests, further complicating the process.
Without more automated and streamlined tools, individuals face unnecessary barriers to effectively controlling their data, undermining the intent of regulations designed to empower them.

In this context, we present a specification to express information regarding rights exercise and management in a machine-interpretable format, using semantic Web standards such as RDF.
In doing so, we begin to address the automation, interoperability and standardisation challenges identified above, towards enabling the development of tools for assisting individuals and organisations in right exercising activities. 
Towards achieving this goal, we reuse and extend the Data Privacy Vocabulary~(DPV)~\cite{panditCreatingVocabularyData2019,pandit2024dpv}, which allows the expression of information related to legislative requirements such as the GDPR.
The developed extension provides a taxonomy to enable representing justifications associated with the non-fulfilment, non-requirement, delays, and exercising of processes, including rights exercising.
Beyond DPV, the contributions in this article also promote the usage of the Open Digital Rights Language (ODRL)~\cite{iannella_odrl_2018}, Data Catalog (DCAT)~\cite{albertoni_dcat_2024}, DCMI Metadata~\cite{dcmi_2020}, and provenance (PROV)~\cite{lebo_prov_2013} ontologies to express information about: 
(1) Linking personal data processing activities to applicable rights,
(2) Providing notices related to said rights,
(3) Documenting the exercise of rights, and
(4) GDPR rights requests.
The resulting specification is being developed in the context of the W3C Data Privacy Vocabularies and Controls Community Group (DPVCG).

@@

Even though the existence of these rights has been brought forward has a huge step for individuals empowerment over their personal data, no standards have been defined for the management of information related to rights exercising.
An exception to this statement might be the `Data Rights Protocol'\footnote{\url{https://datarightsprotocol.org}}, developed by Consumer Reports Innovation Lab\footnote{\url{https://innovation.consumerreports.org}}, which ``defines a web protocol encoding a set of standardized request/response data flows such that Users can exercise Personal Data Rights provided under regulations like the California Consumer Privacy Act, General Data Protection Regulation, and other regulatory or voluntary bases, and receive affirmative responses in standardized formats''\footnote{From \url{https://github.com/consumer-reports-innovation-lab/data-rights-protocol}}.
It does not however cover all GDPR rights and does not promote the usage of standard vocabularies to record information about rights being exercised.
Since semantic technologies promote interoperability, allowing information and its interpretability to be consistently defined across different platforms, improving the discoverability, management, and preservation of resources, we will promote the usage of semantic Web standards for the definition of a specification to express rights exercisement and management.
As such, DPV\footnote{\url{https://w3id.org/dpv}; prefixed as  \texttt{dpv}}\cite{panditCreatingVocabularyData2019,pandit2024dpv} is the main driver of the specification presented in this work, as it is the most comprehensive data protection-related vocabulary, providing a set of taxonomies for the expression of metadata related to the usage of personal data aligned with legal requirements, e.g., the GDPR.
Additionally, the usage of 
ODRL\footnote{\url{http://www.w3.org/ns/odrl/2/} prefixed as  \texttt{odrl}}\cite{iannella_odrl_2018}, 
DCAT\footnote{\url{http://www.w3.org/ns/dcat} prefixed as \texttt{dcat}}\cite{albertoni_dcat_2024}, and 
PROV-O\footnote{\url{http://www.w3.org/ns/prov} prefixed as \texttt{prov}}\cite{lebo_prov_2013}, 
which are Web standards developed and maintained by the W3C, is also promoted to express information about policies, catalogs of resources and activity provenance information, respectively.
The DCMI Metadata Terms standard\footnote{\url{http://purl.org/dc/terms/} prefixed as \texttt{dcterms}}\cite{dcmi_2020}, recommended to be used for metadata-keeping by other other Web standards, including ODRL and DCAT, is also reused.

@@

Based on the GDPR provisions for data subject rights identified in Section~\ref{sec:sota}, the following requirements are addressed in the proposed specification:
\begin{inlineabc}
    \item[(\textit{Section~\ref{sec:applicable-rights}})] information about the existence of rights related to personal data processing activities,
    \item[(\textit{Section~\ref{sec:justifications}})] justifications for the exercising of said rights,
    \item[(\textit{Section~\ref{sec:notices}})] notices related to the fulfilment or non-fulfilment of rights,
    \item[(\textit{Section~\ref{sec:records}})] records of rights-related activities, and
    \item[(\textit{Section~\ref{sec:policies}})] GDPR-related rights requests as machine-executable policies.
\end{inlineabc}
The work is publicly available\footnote{Accessible at  \url{https://besteves4.github.io/dpv-rights-exercising}} and includes examples for each requirement identified above, which are omitted from this article due to size restrictions.

@@

Beyond jurisdiction-dependence, applicable data subject rights also depend on the legal ground used to process personal data.
In the case of the GDPR, DPV's GDPR extension\footnote{\url{https://w3id.org/dpv/legal/eu/gdpr} prefixed as \texttt{eu-gdpr}} contains the terms to represent the rights available under GDPR, as well as a mapping of which rights are applicable based on the used legal basis.
As such, apart from detailing information about what personal data is being processed, how, where, by whom, and for what purpose, a \texttt{dpv:Process} can also be used to indicate applicable rights.
Data controllers can use this approach to express which rights apply, including those beyond GDPR, such as the EU’s fundamental rights and rights outlined in other EU regulations or jurisdictions, i.e., the \texttt{dpv:hasScope} property can be used to indicate applicable rights by jurisdiction.
Listing~\ref{lst:rights} highlights the modeling of a personal data process using DPV, where \texttt{dpv:hasScope} is used to express existing rights under EU's GDPR when data is processed based on consent and its similar UK-based counterpart.

\begin{listing}[ht]
\caption{Associate a process with applicable rights from different jurisdictions}
\label{lst:rights}
\begin{minted}{turtle}
ex:ProcessEmailForServiceProvision a dpv:Process ;
    dpv:hasDataController ex:DataController ;
    dpv:hasPersonalData pd:EmailAddress ;
    dpv:hasProcessing dpv:Use ;
    dpv:hasPurpose dpv:ServiceProvision ;
    dpv:hasScope [
        dpv:hasLegalBasis eu-gdpr:A6-1-a ;
        dpv:hasJurisdiction loc:EU ;
        dpv:hasApplicableLaw legal-eu:law-GDPR ;
        dpv:hasRight eu-gdpr:A7-3, eu-gdpr:A13, eu-gdpr:A14, eu-gdpr:A15, eu-gdpr:A16, 
            eu-gdpr:A17, eu-gdpr:A18, eu-gdpr:A20, eu-gdpr:A22, eu-gdpr:A77 ] ;
    dpv:hasScope [
        dpv:hasLegalBasis ex:GB-GDPR-consent ;
        dpv:hasJurisdiction loc:GB ;
        dpv:hasApplicableLaw legal-gb:law-GDPR, legal-gb:law-DPA ;
        dpv:hasRight ex:RightOfAccess ] . # not complete, other rights exist

@@

As a result of the collection of requirements for GDPR rights exercising, a \texttt{Justification} taxonomy\footnote{\url{https://w3id.org/dpv/justifications} prefixed as \texttt{justifications}}, to provide reasons or explanations related to the fulfilment, non-fulfilment, delay and exercise of rights, was developed.
Examples of justifications include individual's identity not being verifiable or a request being considered overly excessive or burdensome.
Justifications can also explain why certain processes are necessary or being delayed, such as the ground to exercise the right to object or requiring additional information to move forward with a process.
To facilitate the expression of these detailed justifications, this extension introduces concepts that extend DPV's \texttt{Justification} concept.
Moreover, since such concepts can be reused for other data protection-related activities, e.g., data breach reports, the concepts were defined in a generic way so that they can be reused in distinct contexts.

@@

Right exercise notices should inform data subjects about where and how to exercise an active right, what information is needed, or provide updates on a submitted right request.
DPV’s \texttt{isExercisedAt} property can be used to link rights to specific exercise points, such as through right exercise notices, and to connect the required information with a process using DPV’s \texttt{hasProcess} property. DPV's \texttt{hasRecipient}, \texttt{hasStatus}, and \texttt{hasJustification} properties can be employed to update the data subject on the status of their right exercising activity, including any justification for why the right is being exercised.
In addition to the properties mentioned earlier, DCMI's concepts can be reused to indicate temporal and other metadata information related to notices.
Table~\ref{tab:dcterms} summarises the terms reused from DCMI throughout this specification.
The notices should also provide information about the data controller, as well as the entity responsible for implementing the notice.
This could be the data controller or another entity acting on their behalf, represented using \texttt{dpv:hasDataController} and \texttt{dpv:isImplementedByEntity}, respectively. Additionally, \texttt{foaf:page}\footnote{\url{http://xmlns.com/foaf/0.1/} prefixed as \texttt{foaf}} can be used to reference a document, such as a Webpage, that contains the full details of the notice and offers further functionalities.

\begin{table}[ht]
    \centering\fontsize{7}{8}
    \begin{tabular}{|c|c|}
        \hline
        \textbf{\texttt{dcterms} Concepts} & \textbf{Usage} \\ \hline\hline
        \texttt{created},\texttt{issued},\texttt{modified} & Date of creation, issuance or modification of a notice \\ \hline
        \texttt{description} & Description of a notice \\ \hline
        \texttt{identifier} & Identifier of a notice \\ \hline
        \texttt{language} & Language in which the notice is provided \\ \hline
        \texttt{publisher} & Entity responsible for making the notice available \\ \hline
        \texttt{valid} & Date of validity of a notice \\ \hline
    \end{tabular}
    \caption{List of terms reused from the DCMI Metadata Terms vocabulary~\cite{dcmi_2020}}
    \label{tab:dcterms}
\end{table}

@@

\begin{listing}[ht]
\caption{Notice of non-fulfillment related to an exercised right to erasure}
\label{lst:notice}
\begin{minted}{turtle}
ex:RejectRightToErasure a dpv:RightNonFulfilmentNotice ;
    dcterms:issued "2024-09-06"^^xsd:date ;
    dcterms:description "Notice of non-fulfillment related to an exercised right to erasure" ;
    dcterms:identifier "x4ghyun-658393" ;
    dcterms:language "EN" ;
    dcterms:publisher ex:DataController ;
    dpv:hasRight eu-gdpr:A17 ;
    dpv:hasDataController ex:DataController ;
    dpv:isImplementedByEntity ex:DataController ;
    foaf:page <https://example.org/DataController/RejectRightToErasure> ;
    dpv:hasRecipient ex:DataSubject ;
    dpv:hasStatus dpv:RequestUnfulfilled ;
    dpv:hasJustification justifications:FreedomOfExpressionImpaired .

@@

\begin{listing}[ht]
\caption{Record maintained by the data controller for a rights requests}
\label{lst:record}
\begin{minted}{turtle}
ex:catalog-001 a dpv:RightExerciseRecord, dcat:Catalog ;
    dcterms:description "Record maintained by the data controller for a data subject's rights-related requests" ;
    dcterms:publisher ex:DataController ;
    dcterms:created "2023-10-23T08:37:25"^^xsd:dateTime ;
    dcterms:modified "2023-11-03T08:37:25"^^xsd:dateTime ;
    dcat:record ex:DS-record ;
    dcat:catalog ex:request-001, ex:request-002 .

@@

\begin{figure}[ht]
\caption{Lifecycle of Rights Request - modelled as \texttt{Status} concepts in DPV \cite{pandit2024dpv}}
\label{fig:status}
\centering
\fbox{\includegraphics[width=\textwidth]{img/request-status.png}}
\end{figure}

@@

\begin{listing}[ht]
\caption{Right exercise activity where the data controller requires further action from the data subject to be able to proceed with the request}
\label{lst:activity}
\begin{minted}{turtle}
ex:SARRequiresAction a dpv:RightExerciseActivity, prov:Activity ;
    dcterms:description "Data controller requires further action from data subject" ;
    dcterms:created "2023-11-02T16:09:21"^^xsd:dateTime ;
    prov:wasAssociatedWith ex:DataController ;
    dpv:hasDataSubject ex:DataSubject ;
    dpv:hasStatus dpv:RequestRequiresAction ;
    dpv:hasJustification justifications:IdentityVerificationFailure ;
    dpv:hasProcess [
        dpv:hasPersonalData pd:OfficialID ;
        dpv:hasProcessing dpv:MakeAvailable ;
        dpv:hasPurpose dpv:IdentityVerification ;
        dpv:hasRecipientDataController ex:DataController ;
        dpv:isImplementedByEntity ex:DataSubject ] .
\end{minted}
\end{listing}

@@

\begin{listing}[ht]
\caption{Request policy to exercise GDPR right to erasure}
\label{lst:policy}
\begin{minted}{turtle}
ex:delete-request a odrl:Request ;
    odrl:uid "3456-7890-1234-5678-9012"^^xsd:string ;
    dcterms:description "Data subject requests data controller to delete their data." ;
    odrl:obligation ex:DS-delete-data ;
    odrl:obligation ex:DS-delete-notice .

@@

This article provides a machine-interpretable and executable model to represent information regarding data subject rights, concepts to justify their fulfilment or non-fulfilment, notices to communicate about them with other entities, records for auditing and policies for automated execution of responses to rights requests based on the GDPR. Through these, we hope to enable related tools and processes to handle obligations related to rights in an interoperable manner.

We believe our proposal provides a crucially missing technological piece that is required to make effective use of GDPR rights across different services and devices that utilise the Web. The Data Privacy Vocabulary (DPV), which is a central component of our rights management specification, is a comprehensive and growing resource that is being actively developed, has been used by industry and academic projects, and involves inputs from a diverse range of stakeholders. More importantly, the development of DPV -- and the rights management concepts presented here -- involves engagement with legal experts who validate its use as per legal requirements under GDPR, and discussions with Data Protection Officers\footnote{The author Paul Ryan is a functioning DPO for several companies} and implement legal compliance platforms\footnote{The author Georg P. Krog is the chief legal counsel for Signatu -- which uses DPV to provide GDPR compliance solutions} who provide practical grounding for ensuring this work is effective.

To realise the potential of our work, it is important for the rights management specification to be seen as \textit{useful} and \textit{effective} by organisations and legal authorities so as to enable its uptake. To enable its technical adoption, one option is to integrate this work with protocols such as the Data Rights Protocol which creates a single consistent technical signalling mechanism for rights management on the Web. Another option is to utilise existing EU-friendly mechanisms such as the Advanced Data Protection Control\footnote{https://www.dataprotectioncontrol.org/spec/} (ADPC), which have rights management in their scope but lack the technical implementation -- which this work can provide.

Finally, we highlight the importance of this work in developing ecosystems based on the Data Governance Act (DGA), Data Act, and the Data Spaces regulations, where data reuse will be based on technical infrastructures.
% In such ecosystems, having a way to manage rights in a machine-interpretable way is important to reduce the human overload and increase efficiency of legal compliance processes. Our technical specification is forwards compatible as it supports adding more information, workflows, and requirements from these as well as future regulations based on the adaptability of the DPV and the underlying design decisions.
Additionally, the EU Commission has launched an ambitious program\footnote{\url{https://digital-strategy.ec.europa.eu/en/news/eu-digital-identity-4-projects-launched-test-eudi-wallet}} consisting of creating a secure European Digital Identity wallet (EUDI) which will be used by individuals and organisations to assert their identity and access public/private services based on personal data stored in the wallets.
While work on how to store and manage consent in such wallets based on standards has already been explored~\cite{pandit2024implementing}, the question of how such wallets can facilitate rights exercise has not been explored. 
The work presented in this article provides the technical basis for creating a GDPR-compliant by design rights management infrastructure that Data Spaces or EUDI wallets can utilise for individuals and organisations to exercise and manage rights in a consistent standardised and interoperable manner. This not only facilitates GDPR compliance, but also creates a streamlined experience for users and promotes transparency and accountability.

\bibliography{paper}
\bibliographystyle{vancouver}
